---
title: "Run Pipeline"
author: "Mark London"
date: "2026-09-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(autospec)
library(data.table)
```

Run the chunks from top to bottom. Every chunk is safe to run again: nothing
you edit by hand (sign-offs in the plan, code in the suite) is overwritten.
When unsure what to do next, run the `status` chunk. The same steps are in
RStudio's **Addins** menu under *autospec*.

## Settings

Every file path and the round number live in `autospec_config.yaml`. For a new
round, change `round` and `data_file` there. To work on some topics only, list
them under `topics`; `show_topics(config)` lists them.

```{r config}
config <- read_autospec_config("autospec_config.yaml")

# The data's column names, read without loading the data
data_names <- names(haven::read_sav(config$data_file, n_max = 0))
```

## Where am I?

Shows, per topic, how many rows are signed off and still to sign off, any DVs
defined twice that need a decision, how much of the suite is written, and
the chunk to run next.

```{r status}
autospec_status(config, data_names)
```

## 1. Update the plan

Reads the specs, saves the catalogue, and writes `dv_plan.csv`. Run it the first
time, and again whenever the specs change: sign-offs whose spec text has not
changed are kept, the old plan is backed up, and what changed is listed.

DVs defined on more than one sheet with different text are listed in
`QA_duplicate_dvs.csv`. Settle a whole sheet at once by listing the preferred
sheet under `sheet_priority` in the config and running this chunk again.

```{r plan, eval=FALSE}
update_dv_plan(config, data_names)
```

## 1a. Required inputs audit

Writes `inputs.csv` in the output folder, listing every distinct input variable
used by the current plan and whether it was found in the loaded data. When
`name_case` is set to `"lower"`, the variable names are lower-cased before the
check and before writing to CSV.

```{r inputs, eval=FALSE}
plan <- read_dv_plan(config$plan_file)
input_table <- make_required_input_table(plan, data_names = data_names)

if (identical(config$name_case, "lower")) {
  input_table$variable <- tolower(input_table$variable)
  input_table$found_in_data <- tolower(input_table$variable) %in% tolower(data_names)
}

input_table <- input_table[, c("variable", "found_in_data")]
data.table::fwrite(input_table, file.path(config$output_folder, "inputs.csv"))
cli::cli_alert_success("Input audit saved to {.file {file.path(config$output_folder, 'inputs.csv')}}.")
```

## 2. Trial run (optional, any time)

Tries every drafted DV on the data, with nothing signed off. It uses a
throwaway folder, so the plan and the suite are not touched. Change `topics`
to try a few topics; steps in other topics that build their inputs run too.

```{r trial, eval=FALSE}
was_data <- haven::read_sav(config$data_file)

trial <- trial_run(was_data, config, topics = config$topics)
```

## 3. Review the plan

Open `dv_plan.csv` in Excel. For each row, compare `verb`, `args` and
`condition` with `instructions` and read `notes`. When it is right, set `status`
to `reviewed` and fill in `reviewed_by` and `reviewed_on`. Rows marked
`duplicate` or `not_a_derivation` need nothing. Only mark rows reviewed that
have been checked: to see output first, use the trial run.

## 4. Sync the suite with the plan

Writes the suite the first time, and after each batch of sign-offs fills in the
steps that were waiting for them. Steps with code in them, written by hand or
by an earlier sync, are never touched; changed files are backed up in the
suite's `_backup` folder.

```{r sync, eval=FALSE}
sync_dv_suite(
  read_dv_plan(config$plan_file),
  config$suite_folder,
  data_names = data_names,
  name_case = config$name_case,
  topics = config$topics
)
```

Steps still at `derive = NULL` after a sync have no verb that fits: write them
by hand, copying variable names from the step's `inputs`. To try one:

```{r step, eval=FALSE}
run_dv_step(was_data, read_dv_suite(config$suite_folder), "dv_name")
```

## 5. Check and test

Run after every edit. `check_dv_suite()` reads the code without touching the
data; `test_dv_suite()` runs the example tests in the suite's `tests` folder.

```{r check, eval=FALSE}
check_dv_suite(config$suite_folder, data_names = data_names)
test_dv_suite(config$suite_folder)
```

## 6. Build the DVs

Set the pension discount rate in the suite's `settings.R` first. Steps that
fail, are not written yet or are missing an input are reported and skipped.

```{r build, eval=FALSE}
was_data <- haven::read_sav(config$data_file)

was_data <- run_dv_suite(
  was_data,
  config$suite_folder,
  topics = config$topics,
  report_file = config$report_file
)
```

## 7. Save the data with DVs

Saved to `output_data_file` in the config, if it is set.

```{r save, eval=FALSE}
if (nzchar(config$output_data_file)) {
  haven::write_sav(was_data, config$output_data_file)
  cli::cli_alert_success("Data with DVs saved to {config$output_data_file}.")
} else {
  cli::cli_alert_info("output_data_file is blank in the config, so the data was not saved.")
}
```

## 8. Export the DVs as plain R for the team that applies them

Writes plain R scripts, with no autospec in them, to `scripts_folder` in the
config (a checkout of the scripts repo); commit and push from there. The suite
is checked and tested, and the plain R compared DV by DV with the suite on
`was_data`, before anything is written. Export from the whole suite, not a
topic selection.

```{r export, eval=FALSE}
export_dv_scripts(
  config$suite_folder,
  config$scripts_folder,
  was_data,
  title = paste("Spec ", config$round)
)
```
