---
title: "Run Pipeline: Before Review"
author: "Mark London"
date: "2026-09-08"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(autospec)
library(data.table)
read_pipeline_data <- function(path, n_max = NULL) {
	extension <- tolower(tools::file_ext(path))

	if (extension == "csv") {
		return(utils::read.csv(
			path,
			nrows = if (is.null(n_max)) -1L else n_max,
			check.names = FALSE
		))
	}

	if (extension == "sav") {
		if (is.null(n_max)) return(haven::read_sav(path))
		return(haven::read_sav(path, n_max = n_max))
	}

	stop("data_file must be a .csv or .sav file.", call. = FALSE)
}
```
Run the chunks from top to bottom. Every chunk is safe to run again: nothing
you edit by hand (sign-offs in the plan, code in the suite) is overwritten.
When unsure what to do next, run the `status` chunk. The same steps are in
RStudio's Addins menu under autospec.
Settings
Every file path and the round number live in `autospec_config.yaml`. For a new
round, change `round` and `data_file` there. To work on some topics only, list
them under `topics`; `show_topics(config)` lists them.
```{r config}
config <- read_autospec_config("autospec_config.yaml")
# The data's column names, read without loading its rows
data_names <- names(read_pipeline_data(config$data_file, n_max = 0))
```
Where am I?
Shows, per topic, how many rows are signed off and still to sign off, any DVs
defined twice that need a decision, how much of the suite is written, and
the chunk to run next.
```{r status}
autospec_status(config, data_names)
```
1. Update the plan
Reads the specs, saves the catalogue, and writes `dv_plan.csv`. Run it the first
time, and again whenever the specs change: sign-offs whose spec text has not
changed are kept, the old plan is backed up, and what changed is listed.
DVs defined on more than one sheet with different text are listed in
`QA_duplicate_dvs.csv`. Settle a whole sheet at once by listing the preferred
sheet under `sheet_priority` in the config and running this chunk again.
```{r plan, eval=FALSE}
update_dv_plan(config, data_names)
```
1a. Required inputs audit
Writes `inputs.csv` in the output folder, listing every distinct input variable
used by the current plan and whether it was found in the loaded data. When
`name_case` is set to `"lower"`, the variable names are lower-cased before the
check and before writing to CSV.
```{r inputs, eval=FALSE}
plan <- read_dv_plan(config$plan_file)
input_table <- make_required_input_table(plan, data_names = data_names)
if (identical(config$name_case, "lower")) {
input_table$variable <- tolower(input_table$variable)
input_table$found_in_data <- tolower(input_table$variable) %in% tolower(data_names)
}
input_table <- input_table[, c("variable", "found_in_data")]
data.table::fwrite(input_table, file.path(config$output_folder, "inputs.csv"))
cli::cli_alert_success("Input audit saved to {.file {file.path(config$output_folder, 'inputs.csv')}}.")
```
2. Trial run (optional, any time)
Tries every drafted DV on the data, with nothing signed off. It uses a
throwaway folder, so the plan and the suite are not touched. Change `topics`
to try a few topics; steps in other topics that build their inputs run too.
```{r trial, eval=FALSE}
was_data <- read_pipeline_data(config$data_file)
trial <- trial_run(was_data, config, topics = config$topics)
```
3. Review the plan
Open `dv_plan.csv` in Excel. For each row, compare `verb`, `args` and
`condition` with `instructions` and read `notes`. When it is right, set `status`
to `reviewed` and fill in `reviewed_by` and `reviewed_on`. Rows marked
`duplicate` or `not_a_derivation` need nothing. Only mark rows reviewed that
have been checked: to see output first, use the trial run.
Stop here for plan review. Save the reviewed `dv_plan.csv`, then continue with
[`run_pipeline_after_review.Rmd`](run_pipeline_after_review.Rmd).
