---
title: "Run Pipeline: After Review"
author: "Mark London"
date: "2026-10-03"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(autospec)
library(data.table)
read_pipeline_data <- function(path, n_max = NULL) {
  extension <- tolower(tools::file_ext(path))

  if (extension == "csv") {
    return(utils::read.csv(
      path,
      nrows = if (is.null(n_max)) -1L else n_max,
      check.names = FALSE
    ))
  }

  if (extension == "sav") {
    if (is.null(n_max)) return(haven::read_sav(path))
    return(haven::read_sav(path, n_max = n_max))
  }

  stop("data_file must be a .csv or .sav file.", call. = FALSE)
}
```
Continue here after reviewing and saving `dv_plan.csv` from
[`run_pipeline.Rmd`](run_pipeline.Rmd). Each chunk can be rerun; synced code
that has already been written is preserved.
```{r config}
config <- read_autospec_config("autospec_config.yaml")
data_names <- names(read_pipeline_data(config$data_file, n_max = 0))
```
4. Sync the suite with the reviewed plan
Writes the suite the first time, and after each batch of sign-offs fills in the
steps that were waiting for them. Steps with code in them, written by hand or
by an earlier sync, are never touched; changed files are backed up in the
suite's `_backup` folder.
```{r sync, eval=FALSE}
sync_dv_suite(
  read_dv_plan(config$plan_file),
  config$suite_folder,
  data_names = data_names,
  name_case = config$name_case,
  topics = config$topics
)
```
Steps still at `derive = NULL` after a sync have no verb that fits: write them
by hand, copying variable names from the step's `inputs`. To try one:
```{r step, eval=FALSE}
was_data <- read_pipeline_data(config$data_file)
run_dv_step(was_data, read_dv_suite(config$suite_folder), "dv_name")
```
5. Check and test
Run after every edit. `check_dv_suite()` reads the code without touching the
data; `test_dv_suite()` runs the example tests in the suite's `tests` folder.
```{r check, eval=FALSE}
check_dv_suite(config$suite_folder, data_names = data_names)
test_dv_suite(config$suite_folder)
```
6. Build the DVs
Set the pension discount rate in the suite's `settings.R` first. Steps that
fail, are not written yet or are missing an input are reported and skipped.
```{r build, eval=FALSE}
was_data <- read_pipeline_data(config$data_file)
was_data <- run_dv_suite(
  was_data,
  config$suite_folder,
  topics = config$topics,
  report_file = config$report_file
)
```
7. Save the data with DVs
Saved to `output_data_file` in the config, if it is set.
```{r save, eval=FALSE}
if (nzchar(config$output_data_file)) {
  haven::write_sav(was_data, config$output_data_file)
  cli::cli_alert_success("Data with DVs saved to {config$output_data_file}.")
} else {
  cli::cli_alert_info("output_data_file is blank in the config, so the data was not saved.")
}
```
8. Export the DVs as plain R for the team that applies them
Writes plain R scripts, with no autospec in them, to `scripts_folder` in the
config (a checkout of the scripts repo); commit and push from there. The suite
is checked and tested, and the plain R compared DV by DV with the suite on
`was_data`, before anything is written. Export from the whole suite, not a
topic selection.
```{r export, eval=FALSE}
export_dv_scripts(
  config$suite_folder,
  config$scripts_folder,
  was_data,
  title = paste("Spec ", config$round)
)
```
