---
title: "Autospec: Hello World"
author: "Autospec demo"
date: "2026-10-04"
output: html_document
---

## From spec to tested operations

This demo uses the all-verbs test specification and its populated input data. It
runs in an isolated temporary suite and does not modify the working plan or DV
suite.

```{r setup, include=FALSE}
find_project_root <- function(path = getwd()) {
  path <- normalizePath(path, winslash = "/")

  repeat {
    if (file.exists(file.path(path, "autospec.Rproj"))) return(path)
    parent <- dirname(path)
    if (identical(parent, path)) stop("Cannot find autospec.Rproj.", call. = FALSE)
    path <- parent
  }
}

repo_root <- find_project_root()
knitr::opts_knit$set(root.dir = repo_root)
library(autospec)
```

### Human-written spec

Each pseudo-code derivation is catalogued as a requirement. Numbered ranges
expand to individual outputs, each with its own stable UID.

```{r catalogue}
spec_file <- file.path(repo_root, "specs", "TEST_ALL_VERBS.csv")
data_file <- file.path(repo_root, "test_data", "test_input.csv")
input_data <- utils::read.csv(data_file, check.names = FALSE)
catalogue <- make_catalogue(spec_file)
plan <- suppressMessages(build_dv_plan(catalogue, data_names = names(input_data)))

knitr::kable(
  plan[, c("uid", "dv", "verb", "status", "instructions")],
  col.names = c("Requirement UID", "Output", "Operation", "Review status", "Pseudo-code")
)
```

### Input audit

`Found in data` refers strictly to the source CSV. An input can still be
available when an earlier planned operation creates it. The final column shows
whether either source supplies it.

```{r input-audit}
input_audit <- make_required_input_table(plan, data_names = names(input_data))
knitr::kable(
  input_audit[, c("variable", "found_in_data", "provided_by_dvs", "available", "used_by_dvs", "used_by_uids")],
  col.names = c("Input", "Found in CSV", "Produced by", "Available", "Used by outputs", "Consumer UIDs")
)

stopifnot(all(input_audit$available))
```

```{r input-summary, echo=FALSE}
source_count <- sum(input_audit$found_in_data)
produced_count <- sum(!input_audit$found_in_data & nzchar(input_audit$provided_by_dvs))
cat(sprintf(
  "%d required inputs: %d in the CSV, %d produced by suite steps, 0 unavailable.",
  nrow(input_audit), source_count, produced_count
))
```

### Build and run

For this demonstration only, the plan rows are signed off as `DEMO`. The
reviewer and date are written to a temporary plan, then the normal sync, check,
and run steps are used.

```{r pipeline, message=FALSE}
demo_dir <- tempfile("autospec_hello_world")
dir.create(demo_dir)
plan_file <- file.path(demo_dir, "dv_plan.csv")
suite_dir <- file.path(demo_dir, "dv_suite")

plan$status <- "reviewed"
plan$reviewed_by <- "DEMO"
plan$reviewed_on <- as.character(Sys.Date())
write_dv_plan(plan, plan_file)
plan <- read_dv_plan(plan_file)

suppressMessages(sync_dv_suite(
  plan,
  suite_dir,
  data_names = names(input_data),
  name_case = "lower"
))

settings_file <- file.path(suite_dir, "settings.R")
settings <- readLines(settings_file)
settings[settings == "  rate = NULL,"] <- "  rate = 0.05,"
writeLines(settings, settings_file)

suite <- suppressMessages(read_dv_suite(suite_dir))
problems <- suppressMessages(check_dv_suite(suite_dir, data_names = names(input_data)))
stopifnot(!any(problems$severity == "error"))

result <- suppressMessages(run_dv_suite(input_data, suite, stop_on_error = TRUE))
run_report <- attr(result, "dv_suite_report")
```

### UID traceability

The same requirement UID appears in the reviewed plan, generated suite step,
run report, and the unit-test name. Each UID-labelled test checks both the
built value and the UID carried by the suite and run report.

```{r uid-inventory}
report_index <- match(plan$uid, run_report$uid)
operation_inventory <- data.frame(
  requirement_uid = plan$uid,
  output = plan$dv,
  operation = plan$verb,
  suite_uid = vapply(plan$dv, function(dv) suite$steps[[dv]]$uid, character(1L)),
  run_uid = run_report$uid[report_index],
  outcome = run_report$outcome[report_index],
  unit_test_uid = plan$uid,
  unit_test = paste0(
    "Requirement UID ", plan$uid, ": output for ", plan$dv,
    " matches the test data"
  ),
  stringsAsFactors = FALSE
)

stopifnot(
  identical(operation_inventory$requirement_uid, operation_inventory$suite_uid),
  identical(operation_inventory$requirement_uid, operation_inventory$run_uid),
  identical(operation_inventory$requirement_uid, operation_inventory$unit_test_uid),
  all(operation_inventory$outcome == "built")
)

knitr::kable(
  operation_inventory,
  col.names = c("Requirement UID", "Output", "Operation", "Suite UID", "Run UID", "Outcome", "Test UID", "Unit test")
)
```

Run the executable unit tests from the package root in RStudio:

```r
devtools::test(filter = "e2e_all_verbs")
```

The test file is [`tests/testthat/test-e2e_all_verbs.R`](../tests/testthat/test-e2e_all_verbs.R).
