---
title: "Autospec: Hello World"
author: "Autospec demo"
date: "2026-10-04"
output: html_document
---

## From spec to tested operations

This demo uses the all-verbs test specification and its populated input data. It
runs in an isolated temporary suite and does not modify the working plan or DV
suite.

## The idea, plainly

The spec is the requirement; the generated DV is the operation that fulfils
it. Autospec gives each requirement a UID so the plan, generated step, run
result and test can all be tied back to the same line of pseudo-code.

An input listed in the audit is not necessarily supposed to be in the original
CSV. It may be an output of another DV. The audit names that producer and its
UID; readiness then asks whether that producer actually ran successfully.

If two DVs use only source columns, either can run first. A DV that uses their
outputs must wait. In a complete suite, Autospec works out a valid order from
those dependencies; the order of rows in the spec is not the execution order.
When the work is split across stages, the evidence ledger proves which producer
UIDs ran successfully before the next stage.

```{r setup, include=FALSE}
find_project_root <- function(path = getwd()) {
  path <- normalizePath(path, winslash = "/")

  repeat {
    if (file.exists(file.path(path, "autospec.Rproj"))) return(path)
    parent <- dirname(path)
    if (identical(parent, path)) stop("Cannot find autospec.Rproj.", call. = FALSE)
    path <- parent
  }
}

repo_root <- find_project_root()
knitr::opts_knit$set(root.dir = repo_root)
library(autospec)
```

### Human-written spec

Each pseudo-code derivation is catalogued as a requirement. The readable
`REQ-HW-*` IDs are authored beside the derivations in the spec, so reordering
rows does not change requirement identity. Numbered ranges receive deterministic
child IDs for their individual outputs. In each spec block, add an optional
`Requirement UID` (or `UID`) column after `Notes` to assign source IDs; blocks
without it continue to receive generated legacy IDs.

```{r catalogue}
spec_file <- file.path(repo_root, "specs", "TEST_ALL_VERBS.csv")
data_file <- file.path(repo_root, "test_data", "test_input.csv")
input_data <- utils::read.csv(data_file, check.names = FALSE)
catalogue <- make_catalogue(spec_file)
plan <- suppressMessages(build_dv_plan(catalogue, data_names = names(input_data)))

knitr::kable(
  plan[, c("uid", "dv", "verb", "status", "instructions")],
  col.names = c("Requirement UID", "Output", "Operation", "Review status", "Pseudo-code")
)
```

### Input audit

`Found in data` refers strictly to the source CSV. An input can still be
available when an earlier planned operation creates it. The final column shows
whether either source supplies it.

```{r input-audit}
input_audit <- make_required_input_table(plan, data_names = names(input_data))
knitr::kable(
  input_audit[, c("variable", "found_in_data", "provided_by_dvs", "provided_by_uids", "available", "used_by_dvs", "used_by_uids")],
  col.names = c("Input", "Found in CSV", "Produced by", "Producer UIDs", "Available", "Used by outputs", "Consumer UIDs")
)

stopifnot(all(input_audit$available))
```

```{r input-summary, echo=FALSE}
source_count <- sum(input_audit$found_in_data)
produced_count <- sum(!input_audit$found_in_data & nzchar(input_audit$provided_by_dvs))
cat(sprintf(
  "%d required inputs: %d in the CSV, %d produced by suite steps, 0 unavailable.",
  nrow(input_audit), source_count, produced_count
))
```

### Build and run

For this demonstration only, the plan rows are signed off as `DEMO`. The
reviewer and date are written to a temporary plan, then the normal sync, check,
and run steps are used.

```{r pipeline, message=FALSE}
demo_dir <- tempfile("autospec_hello_world")
dir.create(demo_dir)
plan_file <- file.path(demo_dir, "dv_plan.csv")
suite_dir <- file.path(demo_dir, "dv_suite")

plan$status <- "reviewed"
plan$reviewed_by <- "DEMO"
plan$reviewed_on <- as.character(Sys.Date())
write_dv_plan(plan, plan_file)
plan <- read_dv_plan(plan_file)

suppressMessages(sync_dv_suite(
  plan,
  suite_dir,
  data_names = names(input_data),
  name_case = "lower"
))

settings_file <- file.path(suite_dir, "settings.R")
settings <- readLines(settings_file)
settings[settings == "  rate = NULL,"] <- "  rate = 0.05,"
writeLines(settings, settings_file)

suite <- suppressMessages(read_dv_suite(suite_dir))
problems <- suppressMessages(check_dv_suite(suite_dir, data_names = names(input_data)))
stopifnot(!any(problems$severity == "error"))

result <- suppressMessages(run_dv_suite(input_data, suite, stop_on_error = TRUE))
run_report <- attr(result, "dv_suite_report")
```

### UID traceability

The same requirement UID appears in the reviewed plan, generated suite step,
run report, and the unit-test name. Each UID-labelled test checks both the
built value and the UID carried by the suite and run report.

```{r uid-inventory}
report_index <- match(plan$uid, run_report$uid)
operation_inventory <- data.frame(
  requirement_uid = plan$uid,
  output = plan$dv,
  operation = plan$verb,
  suite_uid = vapply(plan$dv, function(dv) suite$steps[[dv]]$uid, character(1L)),
  run_uid = run_report$uid[report_index],
  outcome = run_report$outcome[report_index],
  unit_test_uid = plan$uid,
  unit_test = paste0(
    "Requirement UID ", plan$uid, ": output for ", plan$dv,
    " matches the test data"
  ),
  stringsAsFactors = FALSE
)

stopifnot(
  identical(operation_inventory$requirement_uid, operation_inventory$suite_uid),
  identical(operation_inventory$requirement_uid, operation_inventory$run_uid),
  identical(operation_inventory$requirement_uid, operation_inventory$unit_test_uid),
  all(operation_inventory$outcome == "built")
)

knitr::kable(
  operation_inventory,
  col.names = c("Requirement UID", "Output", "Operation", "Suite UID", "Run UID", "Outcome", "Test UID", "Unit test")
)
```

Run the executable unit tests from the package root in RStudio:

```r
devtools::test(filter = "e2e_all_verbs")
```

The test file is [`tests/testthat/test-e2e_all_verbs.R`](../tests/testthat/test-e2e_all_verbs.R).

## Final totals: readiness by stage

The totals spec says `D = A + B`, `E = B + C` and `F = D + E`, but lists F
first to make the dependency visible. D and E are independent: both use only
source data. F uses their outputs, so it cannot be certified before both
producer UIDs have successful run evidence. This is a dependency rule, not a
rule that D must precede E.

```{r totals-setup}
totals_plan <- suppressMessages(build_dv_plan(
  make_catalogue(file.path(repo_root, "specs", "totals.csv")),
  data_names = names(utils::read.csv(
    file.path(repo_root, "test_data", "totals_input.csv"),
    check.names = FALSE
  ))
))
totals_plan$status <- "reviewed"
totals_plan$reviewed_by <- "DEMO"
totals_plan$reviewed_on <- as.character(Sys.Date())
totals_input <- utils::read.csv(
  file.path(repo_root, "test_data", "totals_input.csv"),
  check.names = FALSE
)
totals_suite_dir <- tempfile("autospec_totals_demo")
totals_evidence_file <- file.path(dirname(totals_suite_dir), "totals_run_evidence.csv")
totals_evidence_id <- "hello-world-totals-snapshot"
suppressMessages(write_dv_suite(
  totals_plan,
  totals_suite_dir,
  data_names = names(totals_input),
  name_case = "spec"
))
totals_suite <- suppressMessages(read_dv_suite(totals_suite_dir))
```

Before D and E run, A/B/C are present, so both producers are ready. F is
blocked, even though its future columns are listed as inputs in the plan: D and
E are not source columns, and their producer requirements have not run yet.

```{r totals-before}
readiness_before <- check_dv_readiness(totals_input, totals_suite)
knitr::kable(readiness_before)
stopifnot(
  all(readiness_before$ready[readiness_before$dv %in% c("D", "E")]),
  !readiness_before$ready[readiness_before$dv == "F"]
)

one_pass <- suppressMessages(run_dv_suite(
  totals_input,
  totals_suite,
  stop_on_error = TRUE
))
knitr::kable(attr(one_pass, "dv_suite_report"))
stopifnot(identical(attr(one_pass, "dv_suite_report")$dv, c("E", "D", "F")))
```

Run D first and save/reload the data. The evidence ledger restores D's run
record after CSV serialization removes dataframe attributes. E is independently
ready and can now run; F remains blocked because E's producer UID is still
pending.

```{r totals-producers}
totals_after_D <- suppressMessages(run_dv_suite(
  totals_input,
  totals_suite,
  dvs = "D",
  with_inputs = FALSE,
  require_ready = TRUE,
  evidence_file = totals_evidence_file,
  evidence_id = totals_evidence_id
))
knitr::kable(attr(totals_after_D, "dv_suite_report"))
totals_after_D_file <- tempfile(fileext = ".csv")
utils::write.csv(totals_after_D, totals_after_D_file, row.names = FALSE)
totals_after_D <- utils::read.csv(totals_after_D_file, check.names = FALSE)

readiness_after_D <- check_dv_readiness(
  totals_after_D,
  totals_suite,
  evidence_file = totals_evidence_file,
  evidence_id = totals_evidence_id
)
knitr::kable(readiness_after_D)
stopifnot(
  readiness_after_D$ready[readiness_after_D$dv == "E"],
  !readiness_after_D$ready[readiness_after_D$dv == "F"],
  identical(readiness_after_D$blocked_by_uids[readiness_after_D$dv == "F"], "REQ-TOTAL-002")
)

totals_after_producers <- suppressMessages(run_dv_suite(
  totals_after_D,
  totals_suite,
  dvs = "E",
  with_inputs = FALSE,
  require_ready = TRUE,
  evidence_file = totals_evidence_file,
  evidence_id = totals_evidence_id
))
knitr::kable(attr(totals_after_producers, "dv_suite_report"))
totals_after_producers_file <- tempfile(fileext = ".csv")
utils::write.csv(totals_after_producers, totals_after_producers_file, row.names = FALSE)
totals_after_producers <- utils::read.csv(totals_after_producers_file, check.names = FALSE)
run_evidence <- utils::read.csv(totals_evidence_file, check.names = FALSE)
knitr::kable(run_evidence[, c("evidence_id", "uid", "dv", "outcome")])
```

After E runs, both producer UIDs are recorded and F becomes ready.
`require_ready = TRUE` enforces that boundary. The evidence ID names this
particular input snapshot; use a new ID if the underlying data changes. The
suite signature also has to match, so editing the producer code or its settings
does not reuse an old success. `trusted_inputs` is an explicit exception for
derived columns verified outside this run.

```{r totals-final}
final_readiness <- check_dv_readiness(
  totals_after_producers,
  totals_suite,
  dvs = "F",
  evidence_file = totals_evidence_file,
  evidence_id = totals_evidence_id
)
knitr::kable(final_readiness)
stopifnot(final_readiness$ready)

totals_result <- suppressMessages(run_dv_suite(
  totals_after_producers,
  totals_suite,
  dvs = "F",
  with_inputs = FALSE,
  require_ready = TRUE,
  evidence_file = totals_evidence_file,
  evidence_id = totals_evidence_id
))
knitr::kable(attr(totals_result, "dv_suite_report"))
knitr::kable(totals_result[, c("A", "B", "C", "D", "E", "F")])
```

Running `F` with `require_ready = TRUE` against the original input is expected
to stop until `REQ-TOTAL-001` and `REQ-TOTAL-002` have succeeded. Set
`trusted_inputs` only when pre-existing derived columns have been independently
verified and are intentionally accepted as upstream results.
