---
title: "Autospec: Hello World"
author: "Autospec demo"
date: "2026-10-04"
output: html_document
---

## From spec to tested operations

This demo uses the all-verbs test specification and its populated input data. It
runs in an isolated temporary suite and does not modify the working plan or DV
suite.

```{r setup, include=FALSE}
find_project_root <- function(path = getwd()) {
  path <- normalizePath(path, winslash = "/")

  repeat {
    if (file.exists(file.path(path, "autospec.Rproj"))) return(path)
    parent <- dirname(path)
    if (identical(parent, path)) stop("Cannot find autospec.Rproj.", call. = FALSE)
    path <- parent
  }
}

repo_root <- find_project_root()
knitr::opts_knit$set(root.dir = repo_root)
library(autospec)
```

### Human-written spec

Each pseudo-code derivation is catalogued as a requirement. The readable
`REQ-HW-*` IDs are authored beside the derivations in the spec, so reordering
rows does not change requirement identity. Numbered ranges receive deterministic
child IDs for their individual outputs. In each spec block, add an optional
`Requirement UID` (or `UID`) column after `Notes` to assign source IDs; blocks
without it continue to receive generated legacy IDs.

```{r catalogue}
spec_file <- file.path(repo_root, "specs", "TEST_ALL_VERBS.csv")
data_file <- file.path(repo_root, "test_data", "test_input.csv")
input_data <- utils::read.csv(data_file, check.names = FALSE)
catalogue <- make_catalogue(spec_file)
plan <- suppressMessages(build_dv_plan(catalogue, data_names = names(input_data)))

knitr::kable(
  plan[, c("uid", "dv", "verb", "status", "instructions")],
  col.names = c("Requirement UID", "Output", "Operation", "Review status", "Pseudo-code")
)
```

### Input audit

`Found in data` refers strictly to the source CSV. An input can still be
available when an earlier planned operation creates it. The final column shows
whether either source supplies it.

```{r input-audit}
input_audit <- make_required_input_table(plan, data_names = names(input_data))
knitr::kable(
  input_audit[, c("variable", "found_in_data", "provided_by_dvs", "provided_by_uids", "available", "used_by_dvs", "used_by_uids")],
  col.names = c("Input", "Found in CSV", "Produced by", "Producer UIDs", "Available", "Used by outputs", "Consumer UIDs")
)

stopifnot(all(input_audit$available))
```

```{r input-summary, echo=FALSE}
source_count <- sum(input_audit$found_in_data)
produced_count <- sum(!input_audit$found_in_data & nzchar(input_audit$provided_by_dvs))
cat(sprintf(
  "%d required inputs: %d in the CSV, %d produced by suite steps, 0 unavailable.",
  nrow(input_audit), source_count, produced_count
))
```

### Build and run

For this demonstration only, the plan rows are signed off as `DEMO`. The
reviewer and date are written to a temporary plan, then the normal sync, check,
and run steps are used.

```{r pipeline, message=FALSE}
demo_dir <- tempfile("autospec_hello_world")
dir.create(demo_dir)
plan_file <- file.path(demo_dir, "dv_plan.csv")
suite_dir <- file.path(demo_dir, "dv_suite")

plan$status <- "reviewed"
plan$reviewed_by <- "DEMO"
plan$reviewed_on <- as.character(Sys.Date())
write_dv_plan(plan, plan_file)
plan <- read_dv_plan(plan_file)

suppressMessages(sync_dv_suite(
  plan,
  suite_dir,
  data_names = names(input_data),
  name_case = "lower"
))

settings_file <- file.path(suite_dir, "settings.R")
settings <- readLines(settings_file)
settings[settings == "  rate = NULL,"] <- "  rate = 0.05,"
writeLines(settings, settings_file)

suite <- suppressMessages(read_dv_suite(suite_dir))
problems <- suppressMessages(check_dv_suite(suite_dir, data_names = names(input_data)))
stopifnot(!any(problems$severity == "error"))

result <- suppressMessages(run_dv_suite(input_data, suite, stop_on_error = TRUE))
run_report <- attr(result, "dv_suite_report")
```

### UID traceability

The same requirement UID appears in the reviewed plan, generated suite step,
run report, and the unit-test name. Each UID-labelled test checks both the
built value and the UID carried by the suite and run report.

```{r uid-inventory}
report_index <- match(plan$uid, run_report$uid)
operation_inventory <- data.frame(
  requirement_uid = plan$uid,
  output = plan$dv,
  operation = plan$verb,
  suite_uid = vapply(plan$dv, function(dv) suite$steps[[dv]]$uid, character(1L)),
  run_uid = run_report$uid[report_index],
  outcome = run_report$outcome[report_index],
  unit_test_uid = plan$uid,
  unit_test = paste0(
    "Requirement UID ", plan$uid, ": output for ", plan$dv,
    " matches the test data"
  ),
  stringsAsFactors = FALSE
)

stopifnot(
  identical(operation_inventory$requirement_uid, operation_inventory$suite_uid),
  identical(operation_inventory$requirement_uid, operation_inventory$run_uid),
  identical(operation_inventory$requirement_uid, operation_inventory$unit_test_uid),
  all(operation_inventory$outcome == "built")
)

knitr::kable(
  operation_inventory,
  col.names = c("Requirement UID", "Output", "Operation", "Suite UID", "Run UID", "Outcome", "Test UID", "Unit test")
)
```

Run the executable unit tests from the package root in RStudio:

```r
devtools::test(filter = "e2e_all_verbs")
```

The test file is [`tests/testthat/test-e2e_all_verbs.R`](../tests/testthat/test-e2e_all_verbs.R).

## Final totals: readiness by stage

The separate totals spec deliberately lists the final output `F` before its
producers `E` and `D`. Readiness is based on requirement UIDs and successful
run evidence, not on row order.

```{r totals-setup}
totals_plan <- suppressMessages(build_dv_plan(
  make_catalogue(file.path(repo_root, "specs", "totals.csv")),
  data_names = names(utils::read.csv(
    file.path(repo_root, "test_data", "totals_input.csv"),
    check.names = FALSE
  ))
))
totals_plan$status <- "reviewed"
totals_plan$reviewed_by <- "DEMO"
totals_plan$reviewed_on <- as.character(Sys.Date())
totals_input <- utils::read.csv(
  file.path(repo_root, "test_data", "totals_input.csv"),
  check.names = FALSE
)
totals_suite_dir <- tempfile("autospec_totals_demo")
totals_evidence_file <- file.path(dirname(totals_suite_dir), "totals_run_evidence.csv")
totals_evidence_id <- "hello-world-totals-snapshot"
suppressMessages(write_dv_suite(
  totals_plan,
  totals_suite_dir,
  data_names = names(totals_input),
  name_case = "spec"
))
totals_suite <- suppressMessages(read_dv_suite(totals_suite_dir))
```

Before D and E run, the source columns make them ready, while F is blocked by
their requirement UIDs.

```{r totals-before}
readiness_before <- check_dv_readiness(totals_input, totals_suite)
knitr::kable(readiness_before)
stopifnot(
  all(readiness_before$ready[readiness_before$dv %in% c("D", "E")]),
  !readiness_before$ready[readiness_before$dv == "F"]
)

one_pass <- suppressMessages(run_dv_suite(
  totals_input,
  totals_suite,
  stop_on_error = TRUE
))
knitr::kable(attr(one_pass, "dv_suite_report"))
stopifnot(identical(attr(one_pass, "dv_suite_report")$dv, c("E", "D", "F")))
```

Run D first and save/reload the data. The evidence ledger restores D's run
record after CSV serialization removes dataframe attributes. E is independently
ready from source columns, but F remains blocked until E's producer UID also
has successful run evidence.

```{r totals-producers}
totals_after_D <- suppressMessages(run_dv_suite(
  totals_input,
  totals_suite,
  dvs = "D",
  with_inputs = FALSE,
  require_ready = TRUE,
  evidence_file = totals_evidence_file,
  evidence_id = totals_evidence_id
))
knitr::kable(attr(totals_after_D, "dv_suite_report"))
totals_after_D_file <- tempfile(fileext = ".csv")
utils::write.csv(totals_after_D, totals_after_D_file, row.names = FALSE)
totals_after_D <- utils::read.csv(totals_after_D_file, check.names = FALSE)

readiness_after_D <- check_dv_readiness(
  totals_after_D,
  totals_suite,
  evidence_file = totals_evidence_file,
  evidence_id = totals_evidence_id
)
knitr::kable(readiness_after_D)
stopifnot(
  readiness_after_D$ready[readiness_after_D$dv == "E"],
  !readiness_after_D$ready[readiness_after_D$dv == "F"],
  identical(readiness_after_D$blocked_by_uids[readiness_after_D$dv == "F"], "REQ-TOTAL-002")
)

totals_after_producers <- suppressMessages(run_dv_suite(
  totals_after_D,
  totals_suite,
  dvs = "E",
  with_inputs = FALSE,
  require_ready = TRUE,
  evidence_file = totals_evidence_file,
  evidence_id = totals_evidence_id
))
knitr::kable(attr(totals_after_producers, "dv_suite_report"))
totals_after_producers_file <- tempfile(fileext = ".csv")
utils::write.csv(totals_after_producers, totals_after_producers_file, row.names = FALSE)
totals_after_producers <- utils::read.csv(totals_after_producers_file, check.names = FALSE)
run_evidence <- utils::read.csv(totals_evidence_file, check.names = FALSE)
knitr::kable(run_evidence[, c("evidence_id", "uid", "dv", "outcome")])
```

Only after both producers have successful run evidence is the final total
ready. `require_ready = TRUE` enforces that boundary.

```{r totals-final}
final_readiness <- check_dv_readiness(
  totals_after_producers,
  totals_suite,
  dvs = "F",
  evidence_file = totals_evidence_file,
  evidence_id = totals_evidence_id
)
knitr::kable(final_readiness)
stopifnot(final_readiness$ready)

totals_result <- suppressMessages(run_dv_suite(
  totals_after_producers,
  totals_suite,
  dvs = "F",
  with_inputs = FALSE,
  require_ready = TRUE,
  evidence_file = totals_evidence_file,
  evidence_id = totals_evidence_id
))
knitr::kable(attr(totals_result, "dv_suite_report"))
knitr::kable(totals_result[, c("A", "B", "C", "D", "E", "F")])
```

Running `F` with `require_ready = TRUE` against the original input is expected
to stop until `REQ-TOTAL-001` and `REQ-TOTAL-002` have succeeded. Set
`trusted_inputs` only when pre-existing derived columns have been independently
verified and are intentionally accepted as upstream results.
