---
title: "From commented SPSS to an audited R function"
author: "autospec"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
find_project_root <- function(path = getwd()) {
  path <- normalizePath(path, winslash = "/")

  repeat {
    if (file.exists(file.path(path, "autospec.Rproj"))) return(path)
    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("Cannot find autospec.Rproj.", call. = FALSE)
    }
    path <- parent
  }
}

repo_root <- find_project_root()
knitr::opts_knit$set(root.dir = repo_root)
devtools::load_all(repo_root, quiet = TRUE)
demo_dir <- tempfile("autospec_spss_demo")
dir.create(demo_dir)
```

## The management example

This short example follows one requirement from its SPSS comment through the
generated specification, reviewed plan and R function, then checks its output.
The UID is authored once in SPSS and follows the requirement through the run
report. The SPSS source lines are retained in the spec and function comments.

The source is [management_demo.sps](../test_SPSS/management_demo.sps):

```{r show-spss, echo=FALSE}
spss_lines <- readLines(
  file.path(repo_root, "test_SPSS", "management_demo.sps"),
  warn = FALSE
)
cat(paste(sprintf("%2d  %s", seq_along(spss_lines), spss_lines), collapse = "\n"))
```

## Create the spec from SPSS

Render this demo from the repository root with:

```r
rmarkdown::render("demos/spss_to_spec.Rmd")
```

The converter writes a CSV spec. We pass that file into the same catalogue and
plan workflow used for hand-written specs.

```{r create-spec}
spss_path <- file.path(repo_root, "test_SPSS", "management_demo.sps")
spec_path <- file.path(demo_dir, "management_demo.csv")
spec_rows <- spss_to_spec(spss_path, output_path = spec_path, quiet = TRUE)

knitr::kable(
  spec_rows[, c(
    "uid", "variable", "derivation", "notes",
    "spss line number(s)", "spss command"
  )],
  col.names = c("Requirement UID", "Output", "Derivation", "Source and notes",
                "SPSS lines", "Command")
)
```

## Draft and review the plan

This requirement is a simple flag. The planner matches it to `dv_flag_if`.
Here the data and sign-off are deliberately small and local to the demo.

```{r plan}
input <- data.frame(var_b04 = c(0, 15, 16, NA_real_))
catalogue <- make_catalogue(spec_path, exclude_sheets = character(), quiet = TRUE)
plan <- suppressMessages(build_dv_plan(catalogue, data_names = names(input)))

knitr::kable(
  plan[, c("uid", "dv", "verb", "inputs", "status", "notes")],
  col.names = c("Requirement UID", "Output", "R operation", "Inputs",
                "Draft status", "Audit trail")
)

# Demo-only sign-off: in a real workflow, review and approve the plan first.
plan$status <- "reviewed"
plan$reviewed_by <- "Demo reviewer"
plan$reviewed_on <- as.character(Sys.Date())
```

## Generate and run the R function

```{r build-and-run}
suite_dir <- file.path(demo_dir, "dv_suite")
suppressMessages(write_dv_suite(
  plan,
  suite_dir,
  data_names = names(input),
  name_case = "lower"
))

topic_file <- list.files(suite_dir, pattern = "[.]R$", full.names = TRUE)
topic_file <- topic_file[basename(topic_file) != "settings.R"]
stopifnot(length(topic_file) == 1L)
generated_function <- readLines(topic_file, warn = FALSE)
cat(paste(generated_function, collapse = "\n"))

result <- suppressMessages(run_dv_suite(input, suite_dir, stop_on_error = TRUE))
stopifnot(isTRUE(all.equal(
  result$var_a02, c(1, 1, 0, NA_real_), check.attributes = FALSE
)))
```

The condition is inclusive: ages 0 and 15 receive 1; age 16 receives 0;
missing age remains missing.

## UID and audit check

The same UID can be checked at the plan, generated step and run-report stages.
The automated test also checks the result and the SPSS line trail:

```{r audit}
run_report <- attr(result, "dv_suite_report")
suite <- suppressMessages(read_dv_suite(suite_dir))
audit <- data.frame(
  requirement_uid = plan$uid,
  output = plan$dv,
  operation = plan$verb,
  suite_uid = vapply(plan$dv, function(dv) suite$steps[[dv]]$uid, character(1L)),
  run_uid = run_report$uid,
  outcome = run_report$outcome,
  source_lines = spec_rows$`spss line number(s)`,
  stringsAsFactors = FALSE
)

stopifnot(
  identical(audit$requirement_uid, audit$suite_uid),
  identical(audit$requirement_uid, audit$run_uid),
  all(audit$outcome == "built")
)
knitr::kable(audit)
```

Run the executable round-trip test from the repository root:

```sh
Rscript -e 'devtools::test(filter = "spss_to_spec")'
```

This demo covers only the simple `DO IF RANGE` / `COMPUTE` / `ELSE` pattern.
It does not attempt to implement the full SPSS language. In particular,
`SELECT var_a TO var_b` depends on variable order in the SPSS dataset, so the
converter records it for review instead of guessing the intervening variables.

```{r cleanup, include=FALSE}
unlink(demo_dir, recursive = TRUE)
```
